Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2016-10-03 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex

Another means of quantifying field data with Streams is the use of contour plots.  Examples of the time averaged contour plots are shown in Figs.~\ref{Gas_and_Gasoline_Contours} and \ref{Foam_Contours} for the natural gas, gasoline, and polyurethane fueled fires.  Each contour line represents the percentage of time that the flame was at a given position. The region under examination, as shown as a black bounding rectangle, is 1400~mm high and 800~mm in width.  The natural gas fire was sampled for 29~s, resulting in 870~images. The area bounded by the contour with the value 1 represents the area that the flame was visible 100~\% of the time sampled. This can be considered the persistent flame region as defined by McCaffrey~\cite{McCaffrey:1979}. The remaining contours are in the intermittent flame region, with the 0.5 contour representing the mean flame height. Outside of the final contour, value 0.0, no visible flame was recorded; hence this contour marks the definitive start of the buoyant plume region.

@@

\section{Experimental Arrangement}
This series of experiments combined the instrumentation used in the non-combustible wall experiments with gypsum wallboard, which has a combustible paper surface. These experiments were conducted in a similar manner as the non-combustible instrumented wall experiments, the only new variable is the painted, gypsum wallboard. These experiments enabled the examination of the ignition/no ignition predictions from the previous experiments.  A free standing, steel frame with a panel of gypsum wallboard attached to the surface was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved toward the wall until the burner was against the wall.  If the results indicated that no burn pattern would result from a given combination of fuel and position, then that fuel and position combination were eliminated from the remainder of the experiments in the study.

@@

Since these experiments were the culmination of the instrumented series of full scale tests, the use of gypsum wallboard was intended to provide some insight into how much fire damage would result. These experiments proved to be challenging in terms of time and effort.  Each panel of gypsum wallboard had to be carefully drilled for thermocouple and heat flux gauge placement. Since there was only one set of heat flux gauges this reduced the number experiments that could be conducted on a given day.  With other project demands, the number of instrumented wallboard experiments that could be conducted under the oxygen consumption calorimeter was reduced due to time constraints. The test matrix is shown in Table~\ref{tab:IWGB_experiments}. There was also a concern that these experiments would have limited value due to potential interference from the water-cooled heat flux gauges and potential physical damage caused by drilling the holes for them.

@@

According to the United States Geological Survey, most gypsum in the United States is used to make wallboards for homes and other structures.  A typical home in the United States contains more than 7 metric tons of gypsum ~\cite{USGS:2015}.  As defined by ASTM, gypsum wallboard is gypsum board designed for use on walls, ceilings, or partitions and that affords a surface suitable to receive decoration~\cite{ASTM_C1396}. The ASTM standard further describes gypsum wallboard as having a noncombustible core, surfaced with paper bonded to the core. By weight, wallboard is composed of more than 85~\% gypsum, less than 10~\% cellulose, less than 3~\% starch and approximately 1~\% of both crystalline silica and fibrous glass~\cite{GA:2010}. Chemically gypsum can be described as calcium sulfate dihydrate, (CaSO$_{4}$ $\bullet$ 2H$_{2}$O) .  By weight pure gypsum is approximately 79~\% calcium sulfate and 21~\% chemically bound water~\cite{Lawson:1977}.

The gypsum wallboard used in these experiments is referred to as `regular' as opposed to gypsum wallboard `Type C' or `Type X' which are used in fire rated assemblies. The gypsum wallboard used in these experiments was purchased from same manufacturer in an effort to have the most similar panels for use between the experiments. Each 1.22~m by 2.44~m by 12.7~mm thick panel weighed approximately 30.5~kg.  The thermal properties of the gypsum wallboard panels are given in the Table~\ref{tab:Gypsum wallboard_Thermal_Properties}.  The first two rows of data were measured from samples of the gypsum wallboard used in these experiments.  The bottom four rows of data are from ~cite\{Gross:1985}.

There have been a number of studies that have examined gypsum wallboard under elevated temperature conditions in terms of it's thermal properties or in terms of its physical change as a result to exposure to heat.  Schroeder conducted a study on the post-fire analysis of construction materials including gypsum.  As gypsum is heated to temperatures above 80~$^\circ$C it begins to change phase and degrade.  Continued heating causes some of the chemically bound water to disassociate and the gypsum changes to a hemihydrate.  This transformation occurs in the range of 98~$^\circ$C to 130~$^\circ$C. With additional heating the residue of the gypsum continues to transform and soften turning into anhydrite, commonly this process is known as calcination.  Schroder exposed samples of gypsum to a range of heat fluxes and temperatures to determine the time for 80~$^\circ$C (dihydrate phase), 200~$^\circ$C (hemihydrate phase), and 500~$^\circ$C (anhydrous phase) isotherms to penetrate to different depths in the gypsum.  The crystal phase and the depth of penetration of the phase change was determined with the use of X~\-ray diffraction~\cite{Schroeder:1999}.  NFPA 921 also discusses calcination and how the investigator can use the visual appearance or probing of the soft dihydrated gypsum to map fire patterns.

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|c|}
		\hline Density & \multicolumn{2}{c|}{Thermal Conductivity}   & \multicolumn{2}{c|}{Specific Heat}  \\
		\hline (kg/m$^3$) & Temperature ($^{\circ}$C) & (W/mK)  & Temperature ($^{\circ}$C)  & (kJ/kg $^{\circ}$C) \\ \hline
		\hline 807  & 24 	& 0.14 	&  		& 1.09 \\
		\hline  	& 100 	& 0.15  &  		&  \\
		\hline 790 	& 20 	& 0.18	& 20 	& 0.90 \\
		\hline  	& 100 	& 0.15 	& 100 	& 0.90 \\
		\hline  	& 200 	& 0.13 	& 200 	& 0.80 \\
		\hline  	& 500 	& 0.15 	& 500	& 0.90 \\
		\hline
		\end{tabular}
		\caption[Thermal properties for 12.7~mm gypsum wallboard]{Thermal properties for 12.7~mm gypsum wallboard}
		\label{tab:Gypsum wallboard_Thermal_Properties}
		\end{table}

After receipt of the gypsum wallboard panels, the panels were laid out on the floor of the NIST Large Fire Research Laboratory, finished paper side up and painted with a roller.  The first coat was a latex primer~\cite{Latex_Primer}.  Once the primer dried a top coat of white latex paint was applied ~\cite{Latex_Paint}.  After the panels dried, they were weighed. The average mass of paint on the finished surface of the gypsum wallboard was 0.33~kg. Then the paint thickness was estimated an ultrasonic coating thickness gauge~\cite{defelsko}.  Over the course of the experiments, random painted panels where checked with the ultra sonic device. The thickness of the paint varied from 0.05~mm to 0.13~mm.  Some of the painted panels were then cut-up into samples for the bench-scale experiments.

@@

Table~\ref{tab:IWGB_experiments} lists the experiments that were conducted for each fuel type.  In this chapter, as in the previous chapter the averaged peak values for temperatures and heat fluxes will presented at varying distances from the wall.  For the gasoline and polyurethane fires, an additional position, 0.08~m from the wall was added.

@@

A complete set of data temperature and heat flux data for the natural gas burner experiments conducted at 0.31~m, 0.15~m, and 0.0~m are provided in Figures~\ref{IWGB_NG_TC_Plume_set},\ref{IWGB_NG_TC_Wall_set},\ref{IWGB_NG_TC_Backside_set},\ref{IWGB_NG_HF_CenterSet}, and \ref{IWGB_NG_HF_EdgeSet}.  The data for the burner positioned 0.31~m from the wall is the average of three replicate experiments.  Note that the data set for the burner position 0.0~m from the wall only has 180~seconds of fire data.  This was the result of a an early shut down of the gas supply.  This decrepancy was not noticed until the data was reviewed.  However at the point there was no opportunity to rerun the experiment.

The temperatures measured along the vertical center axis of the burner provide distinct differences between the flame regions and the buoyant plume region based on the significant differences in temperature see Figure~\ref{IWGB_NG_TC_Plume_set}.  The persistant flame region appears to be recorded at the thermocouple positions 0.2~ through 0.6~m above the burner.  Based on the temperature time histories, the thermocouples positioned 0.8~m through 1.2~m seem to be in the intermittent flame region.  The thermocouples position at or above 1.4~m appear to be in the buoyant plume region.  The fluctuations in the peak temperatures for this ``steady state'' fire are the result of the flame movement away from the thermocouples centered above the burner.  Note that the thermocouple positioned 0.2~m above the burner did not increase in temperature due to a connection problem in the data acquisition system.

@@

The temperatures measured on the front surface of the painted gypsum wallboard are displayed in Figure~\ref{IWGB_NG_TC_Wall_set}.  When the burner was located 0.31~m and 0.15~m away from the wall the temperatures started at ambient and slowly increased  reaching the peak temperatures at 300~seconds when the burner was shut off.  For the case where the burner was in contact with the wall, the two of the thermocouples located in the persistant flame region indicated a decreasing temperature trend after after the paper on the gypsum wallboard had burned away.  The decrease in temperature my have had been the result of material from the burned gysum wallboard shielding otherwise impacting those thermocouples. The thermocouple positioned 0.2~m above the burner was exhibiting the same behavior as in the previous experiment, but the problem was resolved during the experiment.  The restored connection to the data system accounts for the rapid rise in temperature at 60~seconds into the experiment.

@@

Figure~\ref{IWGB_NG_TC_Backside_set} has graphs of the temperature time histories of the themrocouples installed on the back face of the 12.7~mm thick gypsum wallboard.  In each case the temperatures increase while the front face of the gypsum wallboard is exposed to the fire.  As the position of the fire moves closer to the wall the rate of temperature rise on the back increases.  The peak temperatures in each case were approximately 80~$^\circ$C.

@@

The heat flux values from the arrays of sensors mounted flush with the surface of the gypsum wallboard are given in Figures~\ref{IWGB_NG_HF_CenterSet}, and \ref{IWGB_NG_HF_EdgeSet}.  The data from the array centered on the burner and the data from the array located at the right edge of the burner are similar in response and magnitude when the burner is positioned away from the wall.  Similar to the wall surface temperatures the heat flux increase as the burner is moved closer to the wall.  However once the burner is in contact with the wall the flame is moving side to side.  As a result, the peaks in the heat flux alternate between the center array and the edge array, as the flame moves and changes shape.  This movement causes changes in the peak heat flux of approximately 20~kW/m$^2$.  The averaged heat flux values graphed with respect to the burner position from the wall are shown in Figure~\ref{IWGB_NG_HF_CandEAvg}.

@@

Given that the focus of this study is on generating thermal damage on the face of the gypsum wallboard, the temperature and heat flux at the wall surface are important to examine.  The averaged peak temperatures of the wall mounted thermocouples for each of the three different fuels are shown in Figure~\ref{IWGB_Temp_Comp_FrontCenter}. This figure allows the comparison of the averaged peak temperatures at the wall based on the difference in fuel type and based on the burner's distance from the wall.  The graphs in the Figure~\ref{IWGB_Temp_Comp_FrontCenter} show that generally the temperature at the wall surface decreases as the burner moves away from the wall.  The exception to this appears to be at 0.08~m distance from the wall for the gasoline and polyurethane foam experiments.  This position, 1/4 of the width of the burner away from the wall, caused the flame to oscillate toward the wall. As the flame moved toward the wall the flame impact zone on the wall moved around, higher to lower and side to side, which caused the scatter in the averaged data for that fire position.

@@

Figures~\ref{IWGB_HF_Comp_Center} and \ref{IWGB_HF_Comp_Edge} present the averaged peak values from the two heat flux gauge arrays for the natural gas, gasoline and polyurethane foam arrays.  For the gasoline fire experiment at the 0.08~m position the heat flux values from the burner edge array had values that were similar or higher than those measured at the 0.0~m position. As discussed with temperatures above, this was a result of flame movement and the flame briefly contacting the wall near the right edge of the pan more so than seeing the increase at the centerline heat flux array.  A review of the heat release rates for these experiments was made to see if that might have contributed to the higher heat flux, the peak heat release rates for the burner positions at 0.0~m, 0.08~m and 0.15~m were within the average peak heat release rate range of 80~kW~$\pm$~24~kW.  So a change in HRR did not contribute to the increase at the 0.08~m position.

At the 0.15~m position the heat flux exhibits only a small decrease to the 0.2~m and 0.4~m above the burner heat flux sensors in the centerline array.  However, the heat flux sensor positioned at 0.6~m above the burner showed an increase.  As indicated above, the data which goes against the general trend of the data, is a result of the movement of the flame.  For this experiment a sequence of photographs, Figure~\ref{IWGBGAS6_seq}, captured the oscillatory motion of the flame which moved from the vertical centerline toward the wall and back again.  In the process the fire pattern on the wall was generated.

The heat flux data from the polyurethane foam burner experiments also exhibited variations, although less pronounced than those of the gasoline experiments, at the 0.08~m and the 0.15~m position.  The polyurethane foam generated less heat release rate and a shorter flame than the gasoline, so even though the flame was moving toward the wall, impact was not as great.  Again this motion was not a steady motion but an oscillatory one.

@@

Figure~\ref{demarcation} is a photograph of the edge of a fire pattern generated by the natural gas burner.  It will serve as visual definition of the line of demarcation to be used in this study to determine the limits of a fire pattern.  Referring to DeHaan's classifications, there are no surface deposits such as light soot that could be brushed or blown off the gypsum wallboard.  There is a thin band of surface thermal effects in the form of discoloration of the paint, this could be considered an outer bound of the pattern. To the left of the discoloration is clearly an larger area where charring (blackening of the paint/paper), penetration (lifting of the painted/paper residue), and consumption (areas where the paint/paper reside is gone altogether).  The last three classifications or criteria will be used in this study to define the edge of the fire pattern.  Any area where the painted paper surface has been charred, penetrated such that it has lifted away from the gypsum core, or burned away will be considered part of the fire pattern and define the lines of demarcation for this study.

@@

Soot and discoloration alone was not used in this study, since the gasoline and the polyurethane foam generate significant quantities of soot as shown in Figure~\ref{Soot_pattern}.  From the photograph the areas of penetration and consumption can be delineated while the discolored and sooted areas tend to meld together.  Further the focus of the measurements in this study are on thermal damage, therefore the focus is on the burned to not burned line of demarcation.

@@

A summary of the peak average temperatures and heat fluxes at the wall is provided in Table~\ref{tab:IWGB_Summary}. While the values in this summary table differ from the values in Table~\ref{tab:TW_Summary} a closer look reveals that most of the values are not ``significantly '' different.  In other words the range of expanded uncertainty values overlap.  For example at 0.0~m position with natural gas, the average peak temperature was 487~$^\circ$C~$\pm$73~$^\circ$C (15~\%). So the upper bound would be 560~$^\circ$C, which includes the corresponding averaged peak value from the instrumented gypsum wallboard experiments of 539~$^\circ$C.  While all of the values do not have overlapping uncertainties, the point is that none of the averaged peak values are ``that '' different.

@@

Figure~\ref{IWGB_NG_patterns} provides a comparison between the area of discoloration of the paint with the natural gas burner positioned at 0.15~m away from the wall versus the area of consumed and penetrated paint/paper layer which exposed the gypsum underneath.  The peak height of the consumed and penetrated paper, i.e. the burned region, was 0.8~above the burner.  This is consistent with the prediction based on the lower heat flux ignition value found in the literature and the average peak heat flux from the non-combustible was experiments.

@@

Figures~\ref{IWGB_Gas_patterns} and~\ref{IWGB_Gas_0m_patterns} show the progression of thermal damage caused by gasoline fueled burner at 0.15~m, 0.08~m, and 0.0~m away from the gypsum wallboard.  When the burner was positioned at 0.15~m away from the wall, a semi-circular narrow burn line formed with the lowest point approximately 0.1~m above the burner and extending up to the 0.4~m above the burner on the left side.  The formation of the burned area was captured on film and is shown in Figure~\ref{IWGBGAS6_seq}.  The burned region on the gypsum wallboard from exposure to the gasoling fire position 0.08~m away began at 0.2~m above the burner and extended up to 0.6~m. The width of the burned region was approximately one half of the burner width at it's maximum.  Figure~\ref{IWGB_Gas_0m_patterns} shows the fire pattern formed by gas burner when it is tight against the gypsum wallboard.  Two views are shown.  The image on the left shows the wall shortly after the fire went out.  The burned area came be discerned, but a close look is required.  In the photograph on the right, the same piece of gypsum wallboard was lightly brushed off, which revealed a clearer view of the area where the paint and paper were burned off.  The peak height of that fire pattern is just short of 0.8~m. The maxim width is approximately the full width of the burner.

@@

Figures~\ref{IWGB_PUF_patterns} and~\ref{IWGB_PUF_0m_patterns} show the progression of thermal damage caused by the polyurethane fueled burner at 0.15~m, 0.08~m, and 0.0~m away from the instrumented gypsum wallboard.  The burner positioned at 0.15~m from the wall did not transfer enough energy to cause any thermal damage, the only damage to the wall is from soot deposition.  When the burner was located 0.08~m away from the wall, the paint/paper finish over the gypsum was discolored and heavily sooted but the was no consumption or penetration of the paper finish layer.  Three replicate experiments were conducted with the polyurethane foam fueled burner tight against the instrumented gypsum wallboard. The resulting patterns are shown in Figure~\ref{IWGB_PUF_0m_patterns}.  The average height of the burned region of the painted paper finish on the wallboard was 0.3~m above the top of the burner. The maximum width of the burned region was approximately equal to the width of the burner.

@@

The purpose of this chapter is to develop and compare data from bench scale apparatus that could be used to analyze fire damage to gypsum wallboard.
This type of testing might be practical for a fire investigator to use based on the low cost, relative to a full-scale fire test, and the small size of the samples needed for the tests.  In the previous sections of this study, burners were used to examine the repeatability of each fire's characteristics, the amount of heat transfered to a wall as a function of distance and the damage to a section of painted gypsum wallboard.  While the data generated from those full scale laboratory experiments provide valuable insight, it would be better to be able to develop assessments of fire patterns based on data from one or a series of bench scale tests in an effort to validate or improve the existing data in the literature.

>>>>

Another means of quantifying field data with Streams is the use of contour plots.  Examples of the time averaged contour plots are shown in Figures~\ref{Gas_and_Gasoline_Contours} and \ref{Foam_Contours} for the natural gas, gasoline, and polyurethane fueled fires.  Each contour line represents the percentage of time that the flame was at a given position. The region under examination, as shown as a black bounding rectangle, is 1400~mm high and 800~mm in width.  The natural gas fire was sampled for 29~s, resulting in 870~images. The area bounded by the contour with the value 1 represents the area that the flame was visible 100~\% of the time sampled. This can be considered the persistent flame region as defined by McCaffrey~\cite{McCaffrey:1979}. The remaining contours are in the intermittent flame region, with the 0.5 contour representing the mean flame height. Outside of the final contour, value 0.0, no visible flame was recorded; hence this contour marks the definitive start of the buoyant plume region.

@@

\section{Experimental Arrangement}
This series of experiments combined the instrumentation used in the non-combustible wall experiments with gypsum wallboard, which has a combustible paper surface. These experiments were conducted in a manner similar to the non-combustible instrumented wall experiments, the only new variable being the painted, gypsum wallboard. These experiments enabled the examination of the ignition/no ignition predictions from the previous experiments.  A free standing, steel frame with a panel of gypsum wallboard attached to the surface was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved toward the wall until the burner was against the wall.  If the results indicated that no burn pattern would result from a given combination of fuel and position, then that fuel and position combination were eliminated from the remainder of the experiments in the study.

@@

Since these experiments were the culmination of the instrumented series of full scale tests, the use of gypsum wallboard was intended to provide some insight into the amount of fire damage that would result. These experiments proved to be challenging in terms of time and effort.  Each panel of gypsum wallboard had to be carefully drilled for thermocouple and heat flux gauge placement. Since there was only one set of heat flux gauges, this reduced the number experiments that could be conducted on a given day.  With other project demands, the number of instrumented wallboard experiments that could be conducted under the oxygen consumption calorimeter was reduced due to time constraints. The test matrix is shown in Table~\ref{tab:IWGB_experiments}. There was also a concern that these experiments would have limited value due to potential interference from the water-cooled heat flux gauges and potential physical damage caused by drilling the holes for them.

@@

According to the United States Geological Survey, most gypsum in the United States is used to make wallboards for homes and other structures.  A typical home in the United States contains more than 7 metric tons of gypsum ~\cite{USGS:2015}.  As defined by ASTM, gypsum wallboard is gypsum board designed for use on walls, ceilings, or partitions and that affords a surface suitable to receive decoration~\cite{ASTM_C1396}. The ASTM standard further describes gypsum wallboard as having a noncombustible core, surfaced with paper bonded to the core. By weight, wallboard is composed of more than 85~\% gypsum, less than 10~\% cellulose, less than 3~\% starch and approximately 1~\% of both crystalline silica and fibrous glass~\cite{GA:2010}. Chemically, gypsum can be described as calcium sulfate dihydrate, (CaSO$_{4}$ $\bullet$ 2H$_{2}$O) .  By weight, pure gypsum is approximately 79~\% calcium sulfate and 21~\% chemically bound water~\cite{Lawson:1977}.

The gypsum wallboard used in these experiments is referred to as `regular' as opposed to `Type C' or `Type X' gypsum wallboard, which are used in fire rated assemblies. The gypsum wallboard used in these experiments was purchased from the same manufacturer in an effort to have the most similar panels for use between the experiments. Each 1.22~m by 2.44~m by 12.7~mm thick panel weighed approximately 30.5~kg.  The thermal properties of the gypsum wallboard panels are given in the Table~\ref{tab:Gypsum wallboard_Thermal_Properties}.  The first two rows of data were measured from samples of the gypsum wallboard used in these experiments.  The bottom four rows of data are from ~\cite{Gross:1985}.

There have been a number of studies that have examined gypsum wallboard under elevated temperature conditions in terms of either its thermal properties or in terms of its physical change as a result of exposure to heat.  Schroeder conducted a study on the post-fire analysis of construction materials including gypsum.  As gypsum is heated to temperatures above 80~$^\circ$C it begins to change phase and degrade.  Continued heating causes some of the chemically bound water to disassociate and the gypsum changes to a hemihydrate.  This transformation occurs in the range of 98~$^\circ$C to 130~$^\circ$C. With additional heating the residue of the gypsum continues to transform and soften turning into anhydrite, a process commonly known as calcination.  Schroder exposed samples of gypsum to a range of heat fluxes and temperatures to determine the time for 80~$^\circ$C (dihydrate phase), 200~$^\circ$C (hemihydrate phase), and 500~$^\circ$C (anhydrous phase) isotherms to penetrate to different depths in the gypsum.  The crystal phase and the depth of penetration of the phase change was determined with the use of X~\-ray diffraction~\cite{Schroeder:1999}.  NFPA 921 also discusses calcination and how the investigator can use the visual appearance or probing of the soft dihydrated gypsum to map fire patterns.

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|c|}
		\hline Density & \multicolumn{2}{c|}{Thermal Conductivity}   & \multicolumn{2}{c|}{Specific Heat}  \\
		\hline (kg/m$^3$) & Temperature ($^{\circ}$C) & (W/mK)  & Temperature ($^{\circ}$C)  & (kJ/kg $^{\circ}$C) \\ \hline
		\hline 807  & 24 	& 0.14 	& 20	& 1.09 \\
		\hline  	& 100 	& 0.15  &  		&  \\
		\hline 790 	& 20 	& 0.18	& 20 	& 0.90 \\
		\hline  	& 100 	& 0.15 	& 100 	& 0.90 \\
		\hline  	& 200 	& 0.13 	& 200 	& 0.80 \\
		\hline  	& 500 	& 0.15 	& 500	& 0.90 \\
		\hline
		\end{tabular}
		\caption[Thermal properties for 12.7~mm gypsum wallboard]{Thermal properties for 12.7~mm gypsum wallboard}
		\label{tab:Gypsum wallboard_Thermal_Properties}
		\end{table}

After receipt of the gypsum wallboard panels, the panels were laid out on the floor of the NIST Large Fire Research Laboratory, finished paper side up and painted with a roller.  The first coat was a latex primer~\cite{Latex_Primer}.  Once the primer dried a top coat of white latex paint was applied ~\cite{Latex_Paint}.  After the panels dried, they were weighed. The average mass of paint on the finished surface of the gypsum wallboard was 0.33~kg. Then the paint thickness was estimated using an ultrasonic coating thickness gauge~\cite{defelsko}.  Over the course of the experiments, random painted panels where checked with the ultra sonic device. The thickness of the paint varied from 0.05~mm to 0.13~mm.  Some of the painted panels were then cut-up into samples for the bench-scale experiments.

@@

Table~\ref{tab:IWGB_experiments} lists the experiments that were conducted for each fuel type.  In this chapter, as in the previous chapter, the averaged peak values for temperatures and heat fluxes are presented at varying distances from the wall.  For the gasoline and polyurethane fires, an additional position, 0.08~m from the wall, was added.

@@

A complete set of temperature and heat flux data for the natural gas burner experiments conducted at 0.31~m, 0.15~m, and 0.0~m are provided in Figures~\ref{IWGB_NG_TC_Plume_set}, \ref{IWGB_NG_TC_Wall_set}, \ref{IWGB_NG_TC_Backside_set}, \ref{IWGB_NG_HF_CenterSet}, and \ref{IWGB_NG_HF_EdgeSet}.  The data for the burner positioned 0.31~m from the wall is the average of three replicate experiments.  Note that the data set for the burner position 0.0~m from the wall only has 180~seconds of fire data.  This was the result of a an early shut down of the gas supply.  This decrepancy was not noticed until the data was reviewed, at which point there was no opportunity to rerun the experiment.

The temperatures measured along the vertical center axis of the burner provided distinct differences between the flame regions and the buoyant plume region based on the significant differences in temperature, see Figure~\ref{IWGB_NG_TC_Plume_set}.  The persistant or continuous flame region appeared to be recorded at the thermocouple positions 0.2~ through 0.6~m above the burner.  Based on the temperature time histories, the thermocouples positioned 0.8~m through 1.2~m seemed to be in the intermittent flame region.  The thermocouples position at or above 1.4~m appear to be in the buoyant plume region.  The fluctuations in the peak temperatures for this ``steady state'' fire are the result of the flame movement away from the thermocouples centered above the burner.  Note that the thermocouple positioned 0.2~m above the burner did not increase in temperature due to a connection problem in the data acquisition system.

@@

The temperatures measured on the front surface of the painted gypsum wallboard are displayed in Figure~\ref{IWGB_NG_TC_Wall_set}.  When the burner was located 0.31~m and 0.15~m away from the wall, the temperatures started at ambient and slowly increased, reaching the peak temperatures at 300~seconds, then the burner was shut off.  For the case where the burner was in contact with the wall, the two thermocouples located in the persistant flame region indicated a decreasing temperature trend after the paper on the gypsum wallboard had burned away.  The decrease in temperature my have had been the result of material from the burned gysum wallboard shielding otherwise impacting those thermocouples. The thermocouple positioned 0.2~m above the burner was exhibiting the same behavior as in the previous experiment, but the problem was resolved during the experiment.  The restored connection to the data system accounts for the rapid rise in temperature at 60~seconds into the experiment.

@@

Figure~\ref{IWGB_NG_TC_Backside_set} has graphs of the temperature time histories of the thermocouples installed on the back face of the 12.7~mm thick gypsum wallboard.  In each case, the temperatures increase while the front face of the gypsum wallboard is exposed to the fire.  As the position of the fire moves closer to the wall the rate of temperature rise on the back increases.  The peak temperatures in each case were approximately 80~$^\circ$C.

@@

The heat flux values from the arrays of sensors mounted flush with the surface of the gypsum wallboard are given in Figures~\ref{IWGB_NG_HF_CenterSet} and \ref{IWGB_NG_HF_EdgeSet}.  The data from the array centered on the burner and the data from the array located at the right edge of the burner were similar in response and magnitude when the burner is positioned away from the wall.  Similar to the wall surface temperatures, the heat flux increased as the burner was moved closer to the wall.  However, once the burner is in contact with the wall, the flame is moved side to side.  As a result, the peaks in the heat flux alternated between the center array and the edge array, as the flame moved and changed shape.  This movement caused changes in the peak heat flux of approximately 20~kW/m$^2$.  The averaged heat flux values graphed with respect to the burner position from the wall are shown in Figure~\ref{IWGB_NG_HF_CandEAvg}.

@@

Since the focus of this study is on generating thermal damage on the face of the gypsum wallboard, the temperature and heat flux at the wall surface are important to examine.  The averaged peak temperatures of the wall mounted thermocouples for each of the three different fuels are shown in Figure~\ref{IWGB_Temp_Comp_FrontCenter}. This figure allows the comparison of the averaged peak temperatures at the wall based on the difference in fuel type and on the burner's distance from the wall.  The graphs in Figure~\ref{IWGB_Temp_Comp_FrontCenter} show that generally the temperature at the wall surface decreases as the burner moves away from the wall.  The exception to this appears to be at 0.08~m distance from the wall for the gasoline and polyurethane foam experiments.  This position, 1/4 of the width of the burner away from the wall, caused the flame to oscillate toward the wall. As the flame moved toward the wall the flame impact zone on the wall moved around, higher to lower and side to side, which caused the scatter in the averaged data for that fire position.

@@

Figures~\ref{IWGB_HF_Comp_Center} and \ref{IWGB_HF_Comp_Edge} present the averaged peak values from the two heat flux gauge arrays for the natural gas, gasoline and polyurethane foam arrays.  For the gasoline fire experiment at the 0.08~m position the heat flux values from the burner edge array had values that were similar or higher than those measured at the 0.0~m position. As discussed with temperatures above, this was a result of flame movement and the flame briefly contacting the wall near the right edge of the pan more so than seeing the increase at the centerline heat flux array.  A review of the heat release rates for these experiments was made to see if that might have contributed to the higher heat flux, the peak heat release rates for the burner positions at 0.0~m, 0.08~m and 0.15~m were within the average peak heat release rate range of 80~kW~$\pm$~24~kW.  This change in HRR did not contribute to the increase at the 0.08~m position.

At the 0.15~m position the heat flux exhibits only a small decrease to the 0.2~m and 0.4~m above the burner heat flux sensors in the centerline array.  However, the heat flux sensor positioned at 0.6~m above the burner showed an increase.  As indicated above, this data which goes against the general trend of the data, was a result of the movement of the flame.  For this experiment, a sequence of photographs, Figure~\ref{IWGBGAS6_seq}, captured the oscillatory motion of the flame, which moved from the vertical centerline toward the wall and back again.  In the process, the fire pattern on the wall was generated.

The heat flux data from the polyurethane foam burner experiments also exhibited variations, although less pronounced than those of the gasoline experiments, at the 0.08~m and the 0.15~m position.  The polyurethane foam generated a lower heat release rate and a shorter flame than the gasoline, so even though the flame was moving toward the wall, the impact was not as great.  Again, this motion was not a steady motion, but an oscillatory one.

@@

Figure~\ref{demarcation} is a photograph of the edge of a fire pattern generated by the natural gas burner.  It will serve as visual definition of the line of demarcation to be used in this study to determine the limits of a fire pattern.  Referring to DeHaan's classifications, there are no surface deposits such as light soot that could be brushed or blown off the gypsum wallboard.  There is a thin band of surface thermal effects in the form of discoloration of the paint, that could be considered an outer bound of the pattern. To the left of the discoloration is clearly a larger area where charring (blackening of the paint/paper), penetration (lifting of the painted/paper residue), and consumption (areas where the paint/paper residue is gone altogether).  The last three classifications or criteria will be used in this study to define the edge of the fire pattern.  Any area where the painted paper surface has been charred, penetrated such that it has lifted away from the gypsum core, or burned away will be considered part of the fire pattern and define the lines of demarcation for this study.

@@

Soot and discoloration alone were not used in this study, since the gasoline and the polyurethane foam generate significant quantities of soot, as shown in Figure~\ref{Soot_pattern}.  From the photograph, the areas of penetration and consumption can be delineated while the discolored and sooted areas tend to meld together.  Further, the focus of the measurements in this study are on thermal damage.  Therefore the focus is on the ``burned to unburned'' line of demarcation.

@@

A summary of the peak average temperatures and heat fluxes at the wall is provided in Table~\ref{tab:IWGB_Summary}. While the values in this summary table differ from the values in Table~\ref{tab:TW_Summary}, a closer look reveals that most of the values are not ``significantly '' different.  In other words, the range of expanded uncertainty values overlap.  For example, at 0.0~m position with natural gas, the average peak temperature was 487~$^\circ$C~$\pm$73~$^\circ$C (15~\%). So, the upper bound would be 560~$^\circ$C, which includes the corresponding averaged peak value from the instrumented gypsum wallboard experiments of 539~$^\circ$C.  While all of the values do not have overlapping uncertainties, the point is that none of the averaged peak values are substantially different.

@@

Figure~\ref{IWGB_NG_patterns} provides a comparison between the area of discoloration of the paint with the natural gas burner positioned at 0.15~m away from the wall versus the area of consumed and penetrated paint/paper layer which exposed the gypsum underneath when the burner was against the wall. The peak height of the consumed and penetrated paper, i.e. the burned region, was 0.8~above the burner.  This is consistent with the prediction based on the lower heat flux ignition value found in the literature and the average peak heat flux from the non-combustible wall experiments.

@@

Figures~\ref{IWGB_Gas_patterns} and~\ref{IWGB_Gas_0m_patterns} show the progression of thermal damage caused by the gasoline fueled burner at 0.15~m, 0.08~m, and 0.0~m away from the gypsum wallboard.  When the burner was positioned at 0.15~m away from the wall, a semi-circular narrow burn line formed, with the lowest point approximately 0.1~m above the burner and extending up to 0.4~m above the burner on the left side.  The formation of the burned area was captured on film and is shown in Figure~\ref{IWGBGAS6_seq}.  The burned region on the gypsum wallboard from exposure to the gasoline burner position 0.08~m away began at 0.2~m above the burner and extended up to 0.6~m. The width of the burned region was approximately one half of the burner width at its maximum.  Figure~\ref{IWGB_Gas_0m_patterns} shows the fire pattern formed by the gasoline burner when it is tight against the gypsum wallboard.  Two views are shown.  The image on the left shows the wall shortly after the fire went out.  The burned area can be discerned, but a close inspection is required.  In the photograph on the right, the same piece of gypsum wallboard was lightly brushed off, which revealed an improved view of the area where the paint and paper were burned off.  The peak height of that fire pattern is just short of 0.8~m. The maximum width is approximately the full width of the burner.

@@

Figures~\ref{IWGB_PUF_patterns} and~\ref{IWGB_PUF_0m_patterns} show the progression of thermal damage caused by the polyurethane fueled burner at 0.15~m, 0.08~m, and 0.0~m away from the instrumented gypsum wallboard.  The burner positioned at 0.15~m from the wall did not transfer enough energy to cause any thermal damage.  The only damage to the wall is from soot deposition.  When the burner was located 0.08~m away from the wall, the paint/paper finish over the gypsum was discolored and heavily sooted but there was no consumption or penetration of the paper finish layer.  Three replicate experiments were conducted with the polyurethane foam fueled burner placed tight against the instrumented gypsum wallboard. The resulting patterns are shown in Figure~\ref{IWGB_PUF_0m_patterns}.  The average height of the burned region of the painted paper finish on the wallboard was 0.3~m above the top of the burner. The maximum width of the burned region was approximately equal to the width of the burner.

@@

The purpose of this chapter is to develop and compare data from bench scale apparatus that could be used to analyze fire damage to gypsum wallboard.
This type of testing might be practical for a fire investigator to use based on the low cost relative to a full-scale fire test and the small size of the samples needed for the tests.  In the previous sections of this study, burners were used to examine the repeatability of each fire's characteristics, the amount of heat transfered to a wall as a function of distance, and the damage to a section of painted gypsum wallboard.  While the data generated from those full scale laboratory experiments provide valuable insight, it would be better to be able to develop assessments of fire patterns based on data from one or a series of bench scale tests in an effort to validate or improve the existing data in the literature.
